Cultural evolutionary dynamics with prestige bias

### 2022-09-19 ms/ms.tex
continued to edit

\begin{abstract}
Copying role-models can be an efficient method for acquiring knowledge. A common bias when choosing a role-model to copy is success bias: copying whoever appears more successful. This bias depends on the performance of the role-model alone, with no other factors. We propose an additional bias that may be prevalent in cultural transmission: influence bias, in which role-model choice is affected by the number of individuals that have already copied each potential role-model. We combine success and influence bias together to a ``prestige bias'' and analyze its effects on cultural-evolutionary dynamics using mathematical analysis and stochastic simulations. We find analytic approximations to our stochastic model, facilitating further mathematical analysis and reducing the computational complexity of simulations. We validate these approximations using simulations, and demonstrate their robustness to model assumptions.
We also find approximations to the fixation probability and the fixation time of an invading favored (by bias) cultural trait, in both constant and changing environments, which resemble Kimura's approximations for population-genetic models.
These approximations show that success bias effectively plays the part of natural selection, whereas influence bias effectively reduces the population size.	
Influence bias also accelerates the evolutionary dynamics, as can be expected in a \textit{rich-getting-richer} process.
Our model may provide a good description of cultural transmission, 
especially in human societies where social media is popular. 
Further work is needed to test if this model could predict various phenomena in cultural evolution when extended with the effects of selection and innovation.
\end{abstract}

@@

\paragraph{Cultural transmission.}
In cultural transmission, individuals transmit cultural traits (i.e., behavior, beliefs, norms) to one another, typically via learning and demonstrating \citep{transmissionVectorsBook}.
Examples for cultural traits in humans are behavioral patterns, such as personalities and habits, transmitted via both verbally and by observations \citep{cultural_traits}. 
Although cultural transmission is most common in humans, it is also observed in other animals such as chimpanzees \citep{chimpsPrestige, chimpsCopy}, dolphins and whales \citep{dolphins_whales}.
In elephants, \citet{elepahntsRepo} showed that once a matriarch is removed from the group, the group's survival instincts are inferior and that ``the possession of enhanced discriminatory abilities by the oldest individual [matriarch] in a group can influence the social knowledge of the group as a whole.''
By playing audio recordings of African elephants, they showed that groups with a matriarch recognize and react better to hostile or friendly calls than groups without a matriarch.
\citet{fliesPaper} showed that choice of oviposition site in fruit flies is culturally transmitted: inexperienced flies that spent some time with experienced flies chose the same type of oviposition site even without directly observing this behavior. How the information is transmitted is still an open question, but it has been suggested that flies may use olfactory cues like other animals, such as rodents and bees.

\paragraph{Direction of transmission.}
Similar to genetic transmission, culturally transmitted traits can be transmitted from parents to offspring, and their effects of can be physiological rather than behavioral.
For example, parents can "teach" their children to be strong or tall, within some biological limits, by instructing them to maintain a specific diet and engage in physical activity.
Contrary to genetic transmission, cultural transmission can be non-vertical, that is, traits may be transmitted via social learning from non-parental individuals, and even unrelated individuals such as teachers, leaders, media, or any stranger that interacts with the learning individual.
Thus, cultural transmission may combine vertical transmission, where parents transmit to their offspring; oblique transmission, where adults transmit traits to unrelated offspring; and horizontal transmission, where peers from the same age cohort transmit to one another. 
Vertical transmission is also possible in the opposite direction: parents may copy traits from their offspring \citep{transmissionVectorsBook,transmissionVectors}.

\paragraph{Transmission biases.}
In social learning, transmission biases cause a trait to have a disproportionate probability to be transmitted compared to its frequency in the population.
Although more common in cultural transmission, transmission biases do occur in genetic transmission. For example, \textit{wtf genes} in yeast bias their transmission to the gamete by secreting a long life-expectancy poison together with a short life-expectancy antidote, so that a gamete without the gene will perish because the poison will outlive the antidote \citep{wtfGene}.
Importantly, even when a trait is disfavored by natural selection, it may still spread in a population due to transmission biases that are strong enough to overcome selection \citep[Ch. 8 pg. 279]{evolutionBook}.
\citet{cooperation} show that cooperative behavior can evolve in an individual due to horizontal transmission bias even when there is no benefit to it, or when it benefits its competitors.

\paragraph{Success bias.}
\citet{strategiesPaper} have conducted a tournament between learning strategies. Each strategy defines when individuals observe and copy from others, and when they engage in individual learning, in which an individual learns a cultural trait on his own. The best strategies had a high frequency of social learning relative to individual learning, even when the transmission error was almost 50\%. From these results we understand that all the winning strategies were mostly based on success biased social learning, meaning it contributed more to the general success of the population than individual learning. However, all winning strategies included individual learning to some degree, implying that success-biased learning alone isn't the best way to advance, even when success is measured accurately.

\paragraph{Evaluating success.}
\citet[Ch. 5]{evolutionBook} suggest that the evaluation of success can be divided into three groups: \textit{direct bias}, \textit{indirect bias} and \textit{frequency-dependent bias}.
A direct bias occurs when a variation of a trait is more attractive than others, and is evaluated by \textit{directly} testing the variation of the trait.
For example, an individual observing a Ping-Pong match can attempt both of the observed paddle grips to determine which grip is better.
An indirect bias occurs when an individual uses the value of one trait to determine the attractiveness of another, so it \textit{indirectly} evaluates the attractiveness of the role-model.
For example, an observer may copy the paddle grip of the Ping-Pong player who scored more points in the match, thus indirectly evaluating the grip by the points scored.
A frequency-dependent bias occurs when an individual has a probability to copy a variant of the trait that higher or lower than trait's frequency among demonstrators. 
For example, when an individual is 80\% likely to copy the common paddle grip even when only 60\% of the population is using it, it is said to be frequency-biased, or in this case, conformist.
Frequency bias could be negative, i.e., non-conformist bias. 
Conformity and non-conformity are well-known biases in cultural transmission \citep{conformism}, and its effect on cultural evolution have been studied with both models \citep{conformity,anticonformity} and experiments \citep{negativeFrequency}.

\paragraph{Prestige.}
Prestige means having a good reputation or high esteem, therefore does not directly signify success (although it may imply it), making it an indirect bias.
Both \citet[Ch. 8]{evolutionBook} and \citet{complexityPaper} suggest that prestige biases are probably more common in humans than success biases.
\citet[Ch. 8]{evolutionBook} add that maladaptive traits may spread widely in a population if indirect biases are strong enough.
They suggest that such biases could lead to a runaway process caused by a cultural equivalent of sexual selection \citep{sexualSelectionBook}.
On the other hand, \citet{fijian_social_bias} suggest that prestige biases, over generations, can lead to cultural adaptations, and that although prestige can lead to maladaptive traits spreading in the population, it can also accelerate the spread of adaptive traits.
Prestige is often mentioned in the cultural-evolution literature, but seldom modeled.

>>>>

\begin{abstract}
A common bias in cultural transmission is to copy whoever appears more successful. This ``success bias'' only depends on the performance of the role-model.
Here, we propose an additional transmission bias that may be common in cultural transmission: ``influence bias'', in which the choice of a role-model depends on the number of individuals that have already copied from that role-model.
We combine success and influence bias together to form a ``prestige bias'', and analyze its effects on cultural-evolutionary dynamics using mathematical analysis and stochastic simulations.
We find analytic approximations to the stochastic role-model choice process, which facilitate further mathematical analysis and reduces the computational complexity of simulations. We validate these approximations using simulations, and demonstrate their robustness to model assumptions.
We also find approximations to the fixation probability and the fixation time of an invading cultural trait, in both constant and changing environments. 
These approximations are similar to Kimura's approximations for population-genetic models.
These similarities show that success bias effectively plays the role of natural selection, whereas influence bias effectively plays the role of genetic drift.	
Influence bias also accelerates the evolutionary dynamics, as can be expected in a \textit{rich-getting-richer} process.
Our model extends the literature that attempt to describe of cultural transmission, 
especially in human societies where social media is popular and thus influence bias may be expected. 
Further work is needed to test if this model could predict various phenomena in cultural evolution.
\end{abstract}

@@

\paragraph{Cultural transmission.}
In cultural transmission, individuals transmit cultural traits (i.e., behavior, beliefs, norms) to one another, typically via learning and demonstrating \citep{transmissionVectorsBook}.
Examples for cultural traits in humans are behavioral patterns, such as personalities and habits, transmitted via both verbally and by observations \citep{cultural_traits}. 
Although cultural transmission is most common in humans, it is also observed in other animals, e.g., chimpanzees \citep{chimpsPrestige, chimpsCopy}, dolphins and whales \citep{dolphins_whales}, elephants \citep{elepahntsRepo}, and even flies \citep{fliesPaper}.

Cultural transmission may combine vertical transmission, where parents transmit to their offspring; oblique transmission, where adults (teachers, leaders, strangers) transmit traits to unrelated offspring; and horizontal transmission, where peers from the same age cohort transmit to one another. 
Vertical transmission is also possible in the opposite direction: parents may copy traits from their offspring \citep{transmissionVectorsBook,transmissionVectors}.
It has been demonstrated that non-vertical cultural transmission can maintain maladaptive traits, which can be beneficial in changing environments \citep{Ram2018}.

%\paragraph{Transmission biases.}
Cultural transmission biases may cause a trait to have a disproportionate probability to be transmitted compared to its frequency in the population.
%Although more common in cultural transmission, transmission biases do occur in genetic transmission. For example, \textit{wtf genes} in yeast bias their transmission to the gamete by secreting a long life-expectancy poison together with a short life-expectancy antidote, so that a gamete without the gene will perish because the poison will outlive the antidote \citep{wtfGene}.
Importantly, even when a trait is disfavored by natural selection, it may still spread in a population due to transmission biases that are strong enough to overcome selection \citep[Ch. 8 pg. 279]{evolutionBook}. 
For example, \citet{cooperation} have shown that helping behaviors can evolve due to horizontal transmission bias even without any benefit to the recipient, or when the benefit is much larger than the cost.

\paragraph{Success bias.}
% TODO add definition of success bias and some reference to a study demonstrating it in humans or other animals.
Indeed, in a tournament between learning strategies conducted by \citet{strategiesPaper}, most winner included a mixture of success-biased social learning and individual learning, implying that success-biased learning is a good strategy, but that by itself is not enough, even if success is measured accurately.

\citet[Ch. 5]{evolutionBook} have suggested that the evaluation of success can be divided into three groups: \textit{direct bias}, \textit{indirect bias} and \textit{frequency-dependent bias}.
A direct bias occurs when one phenotype is more attractive than other phenotypes, and is evaluated by \textit{directly} testing the trait.
For example, an individual observing a ping-pong match can attempt both of the observed paddle grips to determine which grip is better.
An indirect bias occurs when an individual uses one phenotype to determine the attractiveness of another phenotype.
For example, an observer may copy the paddle grip of the ping-pong player who scored more points in the match, thus indirectly evaluating the grip by the points scored.
A frequency-dependent bias occurs when the probability to copy a phenotype is higher or lower than the frequency of the phenotype among demonstrators. 
For example, when the common paddle grip is used by 60\% of the demonstrators, then if the probability to copy this grip is 80\%, then transmission is under positive frequency bias, also called conformity; if the probability is 40\%, then transmission is under negative frequency bias, or non-conformity \citep{conformism}. 
The effects of conformity and non-conformity on cultural evolution have been studied with both models \citep{conformity,anticonformity} and experiments \citep{negativeFrequency}.

% YOAV HERE
\paragraph{Prestige.}
Prestige means having a good reputation or high esteem, therefore does not directly signify success (although it may imply it), making it an indirect bias.
Both \citet[Ch. 8]{evolutionBook} and \citet{complexityPaper} suggest that prestige biases are probably more common in humans than success biases.
\citet[Ch. 8]{evolutionBook} add that maladaptive traits may spread widely in a population if indirect biases are strong enough.
They suggest that such biases could lead to a runaway process caused by a cultural equivalent of sexual selection \citep{sexualSelectionBook}.
On the other hand, \citet{fijian_social_bias} suggest that prestige biases, over generations, can lead to cultural adaptations, and that although prestige can lead to maladaptive traits spreading in the population, it can also accelerate the spread of adaptive traits.
Prestige is often mentioned in the cultural-evolution literature, but seldom modeled.
